SAT Solving for Vesicle Traffic Systems in Cells

### 2017-07-18 paper/conclusion.tex
added srivas wworkline
some changes

>>>>

In the experiments, we have demonstrated the power of SMT solvers and the value
of careful encoding of the problems into SMT queries.
%
We manage to scale the tool upto the size that makes tool
biologically relevant.
%
However, there are many further search problems that are of interest.
%
For example, are there any k-connected graphs that can not be
annotated into a VTS?
%
This problem induces a quantifier alternation in an encoding.
%
Therefore, a simple call to SAT solver will not work.
%
We are planning to use QBF(quantified Boolean formulas) solvers or develop
iterative search algorithm for such queries.
%

% \ankit{interpret experiments (3-4 lines)}

 \srivas{Future work(quantifier problem etc)}

### 2017-07-18 paper/encoding.tex
some changes

We have encoded several versions of the models of VTSs listed.
%
The cases includes different combinations of constraints on the
activity on nodes and edges.
%

%In this case if the corresponding pairing molecules present on the edge with the molecules , both the molecule and all corresponding molecules gets inhibited/become inactive.

% DIVIDE INTO COLUMNS and TALK ABOUT A/1 or A/0.
%\todo{describe the variations}

\begin{enumerate}[label=\Alph*]
\item Every present molecule is considered to be active.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules. 
\item Activity of molecules on the edges is based on Boolean function of presence of other molecules
\item Activity of molecules both on the edges and nodes is based on Boolean function of presence of other molecules.
\item Activity of molecules on the edges is driven by pairing inhibition.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition.
%\item Version F with additional constraint that every edge is distinct.\ashu{encode this??}
%\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition with every edge is distinct.
%\item Activity of molecules on the nodes is based on Boolean function of presence of other moleculess and on edge by pairing inhibition with self-edges are allowed.
%\item Version F but self-edges are allowed.
\end{enumerate}
%
For the version A, every present molecule is in active state. In the table~\ref{table1}, we list the constraints involved in each version. Constraints V1-V8, R1-R2 and D1-D4 are same for every version but the difference is the constraint deciding the activity of the molecules. One type of restriction is where the activity of the present molecule is dependent on the presence of the other molecules. For example in version B, E, F activity of the molecules on the node is a boolean function of the presence of other molecules on that node; Anb. Similarly for the case of the edge in version C and E; Aeb. In the case of versions F the activity of the molecules on the edges is described by pairing matrix, if all the pairs of the present molecule are present on the same edge than all of them inhibit each other resulting in making all the molecules inactive, we term this process ``pairing inhibition".
%

>>>>

We have encoded the following variants of VTSs.
%
The variants are due to different combinations of constraints on the
activity on nodes and edges.
%
\begin{enumerate}[label=\Alph*]
\item Every present molecule is considered to be active.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules. 
\item Activity of molecules on the edges is based on Boolean function of presence of other molecules
\item Activity of molecules both on the edges and nodes is based on Boolean function of presence of other molecules.
\item Activity of molecules on the edges is driven by pairing inhibition.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition.
\end{enumerate}
%
% For the version A, every present molecule is in active state.
%
In the table~\ref{table1}, we present the constraints involved in each version.
%
The column two of the table shows that the constraints V1-V8, R1-R2, and D1-D4 are present
for every variant.
%
The column three of the table lists the constraints that are different among the variants.
%
%
One of the restriction is where the activity of the present molecule is dependent on the presence of the other molecules.
%
For example in version B, D, F activity of the molecules on the node is a boolean function of the presence of other molecules on that node; Anb.
%
Similarly for the case of the edge in version C and E; Aeb.
%
In the case of versions F the activity of the molecules on the edges
is described by pairing matrix, if all the pairs of the present
molecule are present on the same edge than all of them inhibit each
other resulting in making all the molecules inactive, we term this
process ``pairing inhibition".
%
The constraints for the variants can be given to a SAT solver to find
VTSs that belong to the variants.

### 2017-07-18 paper/experiments.tex
some changes

We have implemented the encoding of the constraints for each version
using the python interface of $\zthree$ in a tool.
%
Our tool allows the user to choose a model and the size of the network
besides other parameters like connectivity and number of parallel
edges.
%
It uses Z3 Python interface to build the needed constraints and
applies $\zthree$ solver on the constraints to find a model ( a
satisfying assignment that respects the constraints).
%
This tool also translates the satisfying model found by $\zthree$ into
a VTS and presents a visual output to the user in form of annotated graph.
%
The output graph will satisfy the underlying rules of the system, so
it always is a valid vesicle transport network.
%
The resultant graph provides an overview of the underlying VTS
where directed labeled edges and nodes provide the
complete information about the system.
%
We also visually report the dropped edges to make the graph
disconnected, which gives information about the connectivity of the
graph.

% Dashed lines are used in the resultant visual network output to
% specify the dropped edges to make the graph disconnected, which gives
% information about the connectivity of the graph.
\ashu{discuss table~\ref{tab:grph}}

@@

In Table~\ref{tab:qf-grabh}, we present the running times for the search of
vesicle traffic systems of sizes 2 to 10 that satisfies the model
variants.
%
The experiments were done on a machine with Intel(R) Core(TM) i3-4030U
CPU @ 1.90GHz processor and 4GB RAM.
%
We have compared our performance with the performance of our earlier 
CBMC based implementation.
%
For example, the formula for variation F, the total number of
compartments ($|N|$) equals to 10, returns in 129.78 minutes (7786.8 secs)
with a SAT result.
%
In comparison, CBMC results in OUT OF MEMORY for $|N|$ greater than 5.
%
Hence with the use of this novel encoding, we are able to scale the
system to a much larger compartmentalized system, especially to
eukaryotic cells with a total number of ten compartments.
%
This additional scalability can provide us leverage to ask for related
open questions that were not previously possible, for example, what is
the minimum number of molecules required to satisfy the vesicular
traffic system in eukaryotes.
%
Furthermore, we experimented with limits of our tools and found
that $\zthree$ was able to solve the constraints up to $\sim{14-18}$ nodes.

>>>>

We have implemented the encodings for each variants
using the python interface of $\zthree$ in a tool.
%
Our tool allows the user to choose a model and the size of the network
besides other parameters like connectivity and number of parallel
edges.
%
It uses Z3 Python interface to build the needed constraints and
applies $\zthree$ solver on the constraints to find a model (
a satisfying assignment that respects the constraints).
%
This tool also translates the satisfying model found by $\zthree$ into
a VTS and presents a visual output to the user in form of annotated graph.
%
% The output graph will satisfy the underlying rules of the system, so
% it always is a valid vesicle transport network.
%
% The resultant graph provides an overview of the underlying VTS
% where directed labeled edges and nodes provide the
% complete information about the system.
%
We also visually report the dropped edges to make the graph
disconnected, which gives information about the connectivity of the
graph.

% Dashed lines are used in the resultant visual network output to
% specify the dropped edges to make the graph disconnected, which gives
% information about the connectivity of the graph.
%\ashu{discuss table~\ref{tab:grph}}

In the table~\ref{tab:grph}, we also present the minimum connectedness needed for
the different variants after applying our tool for sizes from 2 to 10.
%
We haven't found any graph for the variant A with constraints Ann and Aen.
%
Replacing constraint on the node of Ann with Anb (version B) does not affect the outcome.
%
If we allow every present molecule to stay active (Ann) but constraint the edge
by a boolean function (Aeb) the resultant VTS has to be at least 3-connected.
%
If we constrain the activity of both the edge and the nodes with the boolean function (Anb and Aeb), 
the VTS has to be 2-connected.
%
Similarly, the results for the other cases are also presented in the table.

% Again there is no graph for the case where every molecule on the node is active, but the activity on the edge is driven by pairing inhibition (Aep).
% %
% If we replace the constraint on the node to be a boolean function (Anb) with Aep, we have found that 4-connected graphs satisfies all the constraints.

@@

In Table~\ref{tab:qf-grabh}, we present the running times for the search of
VTSs of sizes 2 to 10 that satisfy the variants.
%
The experiments were done on a machine with Intel(R) Core(TM) i3-4030U
CPU @ 1.90GHz processor and 4GB RAM.
%
We have compared our performance with the performance of our earlier 
CBMC based implementation.
%
For example, the formula for variation F, the total number of
compartments ($|N|$) equals to 10, returns in 129.78 minutes (7786.8 secs)
with a SAT result.
%
In comparison, CBMC results in OUT OF MEMORY for $|N|$ greater than 5.
%
``!'' indicate that the constraints were unsatisfiable.
%
Hence with the use of this novel encoding, we are able to scale the
system to a much larger compartmentalized system, especially to
eukaryotic cells with a total number of ten compartments.
%
This additional scalability can provide us leverage to ask for related
open questions that were not previously possible, for example, what is
the minimum number of molecules required to satisfy the vesicular
traffic system in eukaryotes.
%
Furthermore, we experimented with limits of our tools and found
that $\zthree$ was able to solve the constraints up to $\sim{14-18}$ nodes.
